% Experimental setup section — drafted 2026-07-11 for the ICASSP 2026 FedAVOT paper.
% Notation follows the paper (arXiv:2509.14444): p = importance distribution over users,
% q = availability distribution over subsets A_j, T = transport coupling (T 1_M = p,
% 1_N^T T = q), Y[i,j] = T[i,j]/q_j = aggregation weights, theta = model, eta_theta = LR,
% H = local epochs, F(theta) = sum_i p_i f_i(theta).
% NOTE: the repo code names the coupling Y and the weights T — the REVERSE of the paper.
% This file uses the paper's convention.
% Figure references left as \ref{...} placeholders — wire them to the Overleaf labels.

\section{Experimental Setup}
\label{sec:setup}

\subsection{Protocol}
All experiments compare FedAVOT against FedAvg under identical
availability-driven partial participation. There are $N$ users with local
objectives $f_i$, an importance distribution $p$ defining the global
objective $F(\theta)=\sum_{i=1}^{N} p_i f_i(\theta)$, and an availability
distribution $q$ over the $M=\binom{N}{K}$ participating sets
$\mathcal{A}_j \subseteq [N]$ of size $K$. In our experiments $q$ is induced
by per-user availability weights $r \in \Delta^{N-1}$: in every round $t$
the active set $S^t = \mathcal{A}_{j(t)}$ is drawn by sampling $K$ users
without replacement with probabilities proportional to $r$ (via the Gumbel
top-$K$ trick), so that $q_j = \Pr[S^t = \mathcal{A}_j]$; we estimate $q$
from $10^6$ Monte Carlo draws of this sampler ($4\times 10^5$ in the
synthetic experiments). Each active user runs $H=5$ local epochs of
full-batch gradient descent with learning rate $\eta_\theta$ from the
current global model, and the three aggregation rules are updated in
parallel from the same draw $S^t$, so their loss curves are directly
comparable:
\begin{itemize}
  \item \textbf{FedAVOT}:
        $\hat\theta^{t} \leftarrow \sum_{i \in S^t} Y[i, j(t)]\,
        \theta_i^{t-1}$, with weights $Y[i,j] = T[i,j]/q_j$ obtained from
        the masked transport plan $T$ described below;
  \item \textbf{FedAvg ($K$ users)}:
        $\hat\theta^{t} \leftarrow \sum_{i \in S^t} \tfrac{N}{K}\, p_i\,
        \theta_i^{t-1}$;
  \item \textbf{FedAvg (full participation)}:
        $\hat\theta^{t} \leftarrow \sum_{i=1}^{N} p_i\, \theta_i^{t-1}$.
\end{itemize}
The plan $T \in \mathbb{R}^{N \times M}$ is fitted by Sinkhorn scaling to
the marginal constraints $T \mathbf{1}_M = p$ and
$\mathbf{1}_N^{\top} T = q$ with support restricted to the mask
$\mathcal{E} = \{(i,j) : i \in \mathcal{A}_j\}$ (tolerance $10^{-12}$ on
the row marginal, at most $10^3$ iterations for the real-data experiments;
$10^{-10}$ and $3\times 10^3$ for the synthetic ones), and each column of
$Y$ is renormalized to a convex weight vector over $S^t$.

\textbf{Feasibility.} Since $T$ is supported on $\mathcal{E}$, the row
constraint $T\mathbf{1}_M = p$ is achievable only if
\begin{equation}
  p_i \;\le\; \pi_i \;:=\!\! \sum_{j:\, i \in \mathcal{A}_j}\!\! q_j
  \qquad \text{for every user } i,
  \label{eq:feasibility}
\end{equation}
where $\pi_i$ is the inclusion probability of user $i$ under $q$. We
quantify the violation of \eqref{eq:feasibility} as the importance mass
held by infeasible users, $\sum_{i:\, p_i > \pi_i} p_i$, and design the
availability weights $r$ to place each experiment on a chosen side of this
condition.

\textbf{Metrics.} We report the global objective $F(\theta^t)$ with
$\ell$ the squared loss (i.e., the importance-weighted MSE) per round;
final values are averaged over the last $500$ rounds and reported as
mean $\pm$ standard deviation across random seeds.

\subsection{Real data: IMDb-Wiki age regression}
Users are formed by grouping IMDb-Wiki images by celebrity identity; we
take the first $N=100$ identities with at least $30$ training images and
assign each user $n_i = 30$ images. Features are $128$-dimensional face
embeddings extracted with a pretrained ResNet, and the task is linear
regression of (mean-centered) age on the embedding. Hyperparameters:
$K=3$, $4000$ communication rounds ($5000$ for the feasible-availability
variant), $\eta_\theta = 10^{-2}$, $H=5$, $5$ seeds. Two availability
regimes are evaluated against the same cubic importance
$p_i \propto (N{+}1{-}i)^3$:
\begin{itemize}
  \item \emph{Mirrored (infeasible)}: $r_i \propto i^3$, the exact mirror
        image of $p$, so the most important users are almost never
        available; $41/100$ users violate \eqref{eq:feasibility}, holding
        ${\approx}88\%$ of the importance mass;
  \item \emph{Aligned (feasible)}: $r_i \propto N{+}1{-}i$, a linear skew
        in the same direction as $p$; \eqref{eq:feasibility} holds for all
        users ($\max_i p_i/\pi_i = 0.66$) while the cubic-vs-linear shape
        mismatch still requires nontrivial reweighting.
\end{itemize}

\subsection{Controlled synthetic experiment}
To isolate the effect of feasibility we use a linear-regression setting in
which the skew of both distributions is a single knob $\alpha$:
$p_i \propto (N{+}1{-}i)^{\alpha}$ and $r_i \propto i^{\alpha}$, with
$N=50$ and $K=3$. User $i$ holds $30$ samples with standard-normal
features in $d=40$ dimensions, and labels are generated by a user-specific
linear model $\theta_i^{\star} = \bar\theta + h\,\delta_i u +
0.3\,\varepsilon_i$ with additive label noise of standard deviation
$0.1$, where $\delta_i = 2i/(N{-}1)-1$ orders the drift along the
availability axis and $h=3$ sets its strength; that is, the distribution
shift is availability-correlated, so which users participate determines
which optimum is reached. Hyperparameters: $\eta_\theta = 10^{-3}$, $H=5$,
$1500$ communication rounds ($5000$ for the extended feasible run), $3$
seeds. We sweep $\alpha \in \{0.5, 1, 1.5, 2, 3\}$, spanning
${\approx}14\%$ to ${\approx}87\%$ infeasible importance mass; the
feasibility-mechanism diagnostic contrasts $\alpha=0.2$ (fully feasible)
with $\alpha=3.0$.
